Recent advances of malaria parasites detection systems based on mathematical morphology

### 2018-01-25 MDPI_template/Sensors_DiRuberto_rev.tex
Review Jan

\section{Scope of this review}
In this paper we present a review of computer-aided methods oriented to malaria parasites detection and segmentation by mathematical morphology based techniques.
Most of the studies were followed Di Ruberto's work \cite{DiRuberto2002}, which first proposed a system to evaluate parasitaemia in the blood. The system was able to detect the parasites by using an automatic thresholding and morphological operators. A morphological approach to cell segmentation which is more efficient than watershed algorithm \cite{Soille2004} was proposed. Finally, the parasites classification was still based on morphological operators. Since then many systems for computer aided diagnosis of malaria have been proposed. Most of them make use of mathematical morphology to process and analyse malaria-infected peripheral blood cells images. The scope of this paper is to review and analyze the recent works of different researchers in the area of malaria parasite recognition using computer vision which benefit from mathematical morphology.

@@

The proposed automated method in \cite{Elter2011} by Elter \emph{et al.} for parasite detection and identification worked on thin blood film acquired with Giemsa stain. The authors found that the G and B channels of the RGB colour are very good features to identify objects containing chromatin in Giemsa stained blood films, being not only considered highly discriminative but also almost independent of differences in illumination and staining intensity. They transformed the colour input image into a monochrome image \textit{I(x,y)}, that highlights objects containing chromatin: $ I(x,y) = arctan \frac{I_{green}(x,y)}{I_{blue}(x,y)} $. In this work, mathematical morphology has been used with a black top-hat operator to separate MP from both leukocytes and platelets, with a non-flat paraboloid structuring element of radius of 9 and a slope of 1 pixel. It should be taken into account that these fixed parameters might not be suitable for images with different pixel resolutions. The black top-hat operator is followed by a thresholding operation with a fixed threshold, which according to the authors is reliable given the independence of the G and B channels with regard to illumination and staining intensity. However, the authors do not define the value of this fixed threshold on the publication.

@@

In \cite{Ross2006} Ross \emph{et al.} proposed a method which provides a positive or negative diagnosis of malaria and differentiates parasites by species. The segmentation step relies on a thresholding strategy which aims to identify and segment potential parasites and erythrocytes from the image background after a six steps threshold selection. Mathematical morphology has been used for parasite size estimation, erythrocytes reconstruction and cells bigger than erythrocytes removal.

@@

Also in the method proposed in \cite{Somasekar2017} by Somasekar \emph{et al.} the segmentation of the infected parasites is based on thresholding. The segmentation is achieved in two stages by maximizing between-class variance of an original image and consequently by an iterative threshold selection from a stage-one threshold image with suitable stopping criteria. The segmented results are post processed to improve the accuracy of the detection of malaria parasites by morphological operators (erosion and closing).

%Makkapati and Rao proposed two different methodologies. The first one \cite{Makkapati2009} is based on HSV colour space with the purpose of performing a red blood cells (RBCs) segmentation and an identification of RBCs affected from MP with a chromatin dot identification by thresholding the Saturation channel. The RBCs segmentation is achieved by dividing the entire hue range of $360^{\circ}$ into six segments and finding the dominant colour type to be the representative of the background. On the other hand, the second work \cite{Makkapati2011} is devoted to the detection of the parasite stage and identification of the species. It is based on ontological description and uses the morphology of the two basic components of a parasite, namely, chromatin dot(s) and cytoplasm. The authors state that the green component of an image offers maximum separation between the background and RBC pixels. The green component is first segmented to detect the RBCs by choosing an optimal threshold [10]. The pixels belonging to chromatin dots have higher saturation values and are segmented by thresholding the saturation component of the image. However, several artefacts may also be segmented as chromatin
%dots, so a further evaluation is performed to establish if a chromatin dot is valid by identifying whether it is inside a RBC or not.

%The morphology of P.falciparum at gametocyte stage is the key parameter in the segmentation approach proposed in \cite{Kareem2012}. It is based on a modification of the ARR transform. The method can be used as an efficient tool for gametocyte detection for post-treatment malaria diagnosis. It also identifies the presence of any White Blood Cells (WBCs) in the image, and discards other artefacts and non infected cells. It utilizes the information based on structure, colour and geometry of the cells and does not require any segmentation or non-illumination correction techniques that are commonly used for cell detection.

%Fang et al. \cite{Fang2011} used a threshold to distinguish the unhealthy cells from the healthy cells after a malaria detection phase performed with a phase spectrum methodology. This method uses a the Quaternion Fourier Transform (QFT) to obtain the amplitude and phase spectrum for blood smear images. Then, the reconstructed image is obtained using the Inverse Quaternion Fourier transform (IQFT) on a constant amplitude spectrum and the original phase spectrum. The authors concluded that B and G channels from the RGB colour space and I channel from the HSI colour space are the most sensitive channels for MP detection.

%Malihi et al. \cite{Malihi2013} proposed a method for malaria parasite detection in Giemsa-stained blood samples. They used both binary image masks obtained with Otsu method and the edge detection images from Canny method for RBC segmentation, followed by a hole filling process.

%The following strategy has been used:
%\begin{enumerate}
%	\item an algorithm based on run-length encoding has been used for analysing the connected components of the binary masks obtained in segmentation phase;
%	\item preliminary tagging, keeping the label equivalences or connected regions with different labels in a table of local matches. Resolution of the equivalences is, then, performed;
%	\item measurements on the remaining connected components: WBCs and MP region areas are checked and WBCs connected components are removed with a closing operation.
%\end{enumerate}
%

@@

Airwhar \emph{et al.} \cite{Ahirwar2012} based their approach on thresholding and granulometry. The histogram of the complemented, green component has been used and it is said to be a bimodal distribution in all the considered images. Then, both local and global thresholds are used, and the union of the two binary images is chosen as the parasite marker image. A morphological opening filter, using a disk-shaped SE with radius equal to the mean erythrocyte radius less the standard deviation, is applied to the grayscale morphologically filtered green component of the image to remove any objects smaller than an erythrocyte. The morphological gradient is then calculated using a diamond-shaped SE with unity length. The segmentation method is applied to each object in the reconstructed binary image of erythrocytes individually. Those objects that do not exceed the area of a circle with radius equal to the mean erythrocyte radius plus the standard deviation are regarded as being single cells, and are unmodified.
On the other hand, the clumped cells are segmented as follows. First, the intersection of the morphological gradient image and the dilated cell cluster is taken. This image is then transformed to a binary image by thresholding any value greater than zero. A series of morphological operations, namely a closing operation, thinning, and spur removal are then applied to generate a contour of the segmented erythrocytes. The contours are filled, and the segmented mask is again reconstructed with the valid parasite marker image to result in a segmented mask of infected cells.

@@

Soni \emph{et al.} \cite{Soni2011} performed segmentation of erythrocytes by using granulometry as well. The size and eccentricity of the erythrocytes are also required for the calculation of some feature values (as these can be indicative of infection). The shape of the objects (circular erythrocytes) is known a priori, but the image must be analysed to determine the size distribution of objects in the image and to find the average eccentricity of erythrocytes present.
Gray-scale granulometries based on opening with disk-shaped elements are then used. Non flat disk-shaped structural element are applied to enhance the roundness and compactness of the red blood cells and flat disk-shaped structural element applied to segment overlapping cells. The object to be segmented differs greatly in contrast from the background image. Changes in contrast can be detected by operators that calculate the gradient of an image. The gradient image can be computed and a threshold can be applied to create a binary mask containing the segmented cell. The binary gradient mask is dilated using a vertical structuring element followed by a horizontal structuring element. The cell of interest has been successfully segmented, but it is not the only object that has been found. Any objects that are connected to the border of the image can be removed.

@@

%Zou et al. \cite{Zou2010} based their method on the consideration that RBCs have a circular shape. They propose an improved version of the Circle Hough Transform (CHT) for RBCs detection. The authors state that overlapping and oval cells can be detected as well as non-clumped RBCs and affirm that the green channel of the RGB colour space permits to better distinguish RBCs. However, some RBCs may not have circular shape and it could affect the detection results.

To conclude, many systems for computer aided diagnosis of malaria disease made use of mathematical operations in order to smoothen the boundary of the regions obtained from the segmentation process.

@@

%

@@

Malaria parasites may be discriminated according to two different strategies: by segmenting the whole erythrocyte from the blood smear image on the basis of which malaria infection is detected, otherwise by segmenting chromatin dot or parasite infection region for characterizing parasite infection stages based on some extracted target features.
In general, thresholding-based approach is still widely used for segmentation purposes. In particular, a lot of authors affirm that Otsu thresholding suffers from limitations when textural variation is high, while histogram thresholding can not deal sufficiently good in identifying valley regions in case of unimodal histograms. However, such a simple and fast approach can greatly benefit from mathematical morphology as recent studies demonstrate (\cite{Ahirwar2012}, \cite{Anggraini2011}, \cite{Arco2014}, \cite{DiRuberto2002}, \cite{Elter2011}, \cite{Ghosh2011},  \cite{Kareem2011}, \cite{Mushabe2013},  \cite{Rosado2017}, \cite{Ross2006}, \cite{Savkare2011b}, \cite{Savkare2015}, \cite{Somasekar2017}, \cite{Tek2010}).

@@

Other authors (\cite{Ahirwar2012}, \cite{DiRuberto2002}, \cite{Khan2011}, \cite{Mushabe2013}, \cite{Ross2006}, \cite{Soni2011}, \cite{Tek2010}) employed granulometry and stated that it is very effective to segment cells with regular size.

The analysed works performed classification phase for different purposes. The majority of them aimed to distinguish among two classes only, malaria infected and noninfected RBCs, or to detect and count parasites in a malaria blood image (\cite{Anggraini2011}, \cite{Arco2014}, \cite{Das2011}, \cite{Dave2017}, \cite{Das2015}, \cite{DiRuberto2002}, \cite{Elter2011}, \cite{Ghosh2011}, \cite{Kareem2012}, \cite{Khan2014}, \cite{Malihi2013}, \cite{Mushabe2013}, \cite{Oliveira2017}, \cite{Savkare2011b}, \cite{Savkare2011a}, \cite{Somasekar2011}, \cite{Somasekar2015}, \cite{Somasekar2017}, \cite{Soni2011}, \cite{Sulist2015}, \cite{Tek2006}).

More complex classification strategies aimed to classify parasites into different classes, i.e. different human parasites species (\cite{Ahirwar2012}, \cite{Das2013}, \cite{Das2014}, \cite{Khan2011}, \cite{Tek2010}),
and/or different parasites life stages (\cite{Anggraini2011}, \cite{Das2013}, \cite{Das2014}, \cite{DiRuberto2002}, \cite{Diaz2009}, \cite{Tek2010}).

A summary of analysed methods is shown in Table 1.

@@

It is worth noticing that cells colours and the colour contrast between cells and background can vary so often according to the different, existing staining techniques, thickness of smear, microscope illumination and microscope's image acquisition procedure, as shown in fig. \ref{fig:images_types}. A standardization of the procedure should be really useful to avoid superfluous differences in similar images' features and to have fair comparisons among the several proposed methods. The main efforts towards the realization of a fully automatic blood cells segmentation and classification system cannot leave this aspect out.

Mathematical morphology techniques have been widely used for image processing purposes. Among the application fields, it has been applied for fingerprint feature extraction, recognition of handwritten digits, license plate detection, border extraction, denoising using morphological filters, text extraction and detection of imperfection in printed circuit boards \cite{Kaur2013}. Apart from this kind of fields, mathematical morphology has been employed successfully in biomedical image analysis, especially in preprocessing and segmentation techniques.

@@

%=====================================
% References, variant B: external bibliography
%=====================================
\externalbibliography{yes}
\bibliography{bibliography}

>>>>

\section{Scope of this review}
In this paper we present a review of computer-aided methods oriented to malaria parasites detection and segmentation by mathematical morphology based techniques.
Most of the studies were followed Di Ruberto's work \cite{DiRuberto2002}, which first proposed a system to evaluate parasitaemia in the blood. The system was able to detect the parasites by using an automatic thresholding and morphological operators. A morphological approach to cell segmentation which is more efficient than watershed algorithm \cite{Soille2004} was proposed. Finally, the parasites classification was still based on morphological operators. Since then many systems for computer aided diagnosis of malaria have been proposed. Most of them make use of mathematical morphology to process and analyse malaria-infected peripheral blood cells images. The scope of this paper is to review and analyze the recent works of different researchers in the area of malaria parasite recognition using computer vision which benefit from mathematical morphology. Only few reviews exist in literature about microscopic image processing for malaria parasites. However, they are not focused on MM techniques as they analyze generic computer vision systems for malaria diagnosis, as in \cite{Rosado2017}, \cite{Das2015} or \cite{Tek2009}. Also, newer and promising approaches, addressing for example the problem of malaria diagnosis in remote areas \cite{Oliveira2017}, \cite{Rosado2017}, or improving significantly both the detection and the classification performances \cite{Dave2017}, \cite{Devi2017}, \cite{Somasekar2017} and \cite{Sulist2015}, have not been considered in the previous reviews.

@@

The automated method proposed in \cite{Elter2011} by Elter \emph{et al.} for parasite detection and identification worked on thin blood film acquired with Giemsa stain. The authors found that the G and B channels of the RGB colour are very good features to identify objects containing chromatin in Giemsa stained blood films, being not only considered highly discriminative but also almost independent of differences in illumination and staining intensity. They transformed the colour input image into a monochrome image \textit{I(x,y)}, that highlights objects containing chromatin: $ I(x,y) = arctan \frac{I_{green}(x,y)}{I_{blue}(x,y)} $. In this work, mathematical morphology has been used with a black top-hat operator to separate MP from both leukocytes and platelets, with a non-flat paraboloid structuring element of radius of 9 and a slope of 1 pixel. It should be taken into account that these fixed parameters might not be suitable for images with different pixel resolutions. The black top-hat operator is followed by a thresholding operation with a fixed threshold, which according to the authors is reliable given the independence of the G and B channels with regard to illumination and staining intensity. However, the authors do not define the value of this fixed threshold on the publication.

@@

In \cite{Ross2006} Ross \emph{et al.} proposed a method which provides a positive or negative diagnosis of malaria and differentiates parasites by species. The segmentation step relies on a six steps thresholding selection strategy. It aims to identify and segment potential parasites and erythrocytes from background. Mathematical morphology has been used in several key steps of the procedure. Hole filling is used in the first step in order to fill RBCs' binary masks obtained from a first thresholding. Afterwards, step 4 employs RBCs' morphological reconstruction with parasites' mask, found in step 2, for identifying infected cells. In step 5, a morphological opening filter, using a disk-shaped SE with radius equal to the mean erythrocyte radius less the standard deviation, is applied to the grey-scale, morphologically filtered, green component in order to remove any objects smaller than an erythrocyte. The morphological gradient - difference between a dilation and erosion of the image - is then calculated using a diamond-shaped SE with unity length.
Finally, in step 6, the intersection of morphological gradient image and the dilated cell cluster is calculated. This image is then transformed to a binary image by thresholding any value greater than zero. A series of morphological operations, namely a closing operation, thinning, and spur-removal are then applied to generate a contour of the segmented erythrocytes. Contours are filled, and the segmented mask is again reconstructed with the valid parasite marker image to result in a segmented mask of infected cells. RBCs and parasites masks are consequently ready for next generation step.

@@

Also in the method proposed in \cite{Somasekar2017} by Somasekar \emph{et al.} the segmentation of the infected parasites is based on thresholding. It is achieved in two stages by maximizing between-class variance of an original image and consequently by an iterative threshold selection from a stage-one threshold image with suitable stopping criteria. The segmented results are post processed to improve the accuracy of malaria parasites detection by morphological operators (erosion and closing).

@@

%Airwhar \emph{et al.} \cite{Ahirwar2012} based their approach on thresholding and granulometry. The histogram of the complemented, green component has been used and it is said to be a bimodal distribution in all the considered images. Then, both local and global thresholds are used, and the union of the two binary images is chosen as the parasite marker image. A morphological opening filter, using a disk-shaped SE with radius equal to the mean erythrocyte radius less the standard deviation, is applied to the grayscale morphologically filtered green component of the image to remove any objects smaller than an erythrocyte. The morphological gradient is then calculated using a diamond-shaped SE with unity length. The segmentation method is applied to each object in the reconstructed binary image of erythrocytes individually. Those objects that do not exceed the area of a circle with radius equal to the mean erythrocyte radius plus the standard deviation are regarded as being single cells, and are unmodified.
%On the other hand, the clumped cells are segmented as follows. First, the intersection of the morphological gradient image and the dilated cell cluster is taken. This image is then transformed to a binary image by thresholding any value greater than zero. A series of morphological operations, namely a closing operation, thinning, and spur removal are then applied to generate a contour of the segmented erythrocytes. The contours are filled, and the segmented mask is again reconstructed with the valid parasite marker image to result in a segmented mask of infected cells.

@@

%Soni \emph{et al.} \cite{Soni2011} performed segmentation of erythrocytes by using granulometry as well. The size and eccentricity of the erythrocytes are also required for the calculation of some feature values (as these can be indicative of infection). The shape of the objects (circular erythrocytes) is known a priori, but the image must be analysed to determine the size distribution of objects in the image and to find the average eccentricity of erythrocytes present.
%Gray-scale granulometries based on opening with disk-shaped elements are then used. Non flat disk-shaped structural element are applied to enhance the roundness and compactness of the red blood cells and flat disk-shaped structural element applied to segment overlapping cells. The object to be segmented differs greatly in contrast from the background image. Changes in contrast can be detected by operators that calculate the gradient of an image. The gradient image can be computed and a threshold can be applied to create a binary mask containing the segmented cell. The binary gradient mask is dilated using a vertical structuring element followed by a horizontal structuring element. The cell of interest has been successfully segmented, but it is not the only object that has been found. Any objects that are connected to the border of the image can be removed.

@@

To sum up, the analysed systems use mathematical morphology methods during or immediately ensuing segmentation step for the following purposes:

\begin{itemize}
	\item size evaluation of regions obtained from segmentation
	\item image cleaning and artefacts removal
	\item RBCs and parasites segmentation or separation
	\item improving parasites detection
	\item minimization of watershed over-segmentation.
\end{itemize}

@@

@@

Malaria parasites may be discriminated according to two different strategies: by segmenting the whole erythrocyte from the blood smear image on the basis of which malaria infection is detected, otherwise by segmenting chromatin dot or parasite infection region for characterizing parasite infection stages based on some extracted target features.
In general, thresholding-based approach is still widely used for segmentation purposes. In particular, a lot of authors affirm that Otsu thresholding suffers from limitations when textural variation is high, while histogram thresholding can not deal sufficiently good in identifying valley regions in case of unimodal histograms. However, such a simple and fast approach can greatly benefit from mathematical morphology as recent studies demonstrate, \cite{Anggraini2011}, \cite{Arco2014}, \cite{DiRuberto2002}, \cite{Elter2011}, \cite{Ghosh2011},  \cite{Kareem2011}, \cite{Mushabe2013},  \cite{Rosado2017}, \cite{Ross2006}, \cite{Savkare2011b}, \cite{Savkare2015}, \cite{Somasekar2017}, \cite{Tek2010}).

@@

Other authors, \cite{DiRuberto2002}, \cite{Khan2011}, \cite{Mushabe2013}, \cite{Ross2006}, \cite{Tek2010}) employed granulometry and stated that it is very effective to segment cells with regular size.

The analysed works performed classification phase for different purposes. The majority of them aimed to distinguish among two classes only, malaria infected and noninfected RBCs, or to detect and count parasites in a malaria blood image (\cite{Anggraini2011}, \cite{Arco2014}, \cite{Das2011}, \cite{Dave2017}, \cite{Das2015}, \cite{DiRuberto2002}, \cite{Elter2011}, \cite{Ghosh2011}, \cite{Kareem2012}, \cite{Khan2014}, \cite{Malihi2013}, \cite{Mushabe2013}, \cite{Oliveira2017}, \cite{Savkare2011b}, \cite{Savkare2011a}, \cite{Somasekar2011}, \cite{Somasekar2015}, \cite{Somasekar2017}, \cite{Sulist2015}, \cite{Tek2006}).

More complex classification strategies aimed to classify parasites into different classes, i.e. different human parasites species \cite{Das2013}, \cite{Das2014}, \cite{Khan2011}, \cite{Tek2010}),
and/or different parasites life stages (\cite{Anggraini2011}, \cite{Das2013}, \cite{Das2014}, \cite{DiRuberto2002}, \cite{Diaz2009}, \cite{Tek2010}).

A summary of analysed methods is shown in Table 1. For each approach we report the processing phase where MM is used (preprocessing and/or segmentation), which operators are applied, which type of morphology (grayscale or binary), and shape and size of the used SE, if described. We detail the kind of classification, i.e. if a method addresses the detection problem (infected and non infected) only, if it faces the labelling of different parasites (\emph{P.falciparum}, \emph{P.ovale}, \emph{P.malariae} or \emph{P.vivax}) and, in some cases, the different stages of life (ring, trophozoite, schizont, gametocyte). Finally, the measures used for evaluating the system performance are reported, if present.

@@

It is worth noticing that cells colours and the colour contrast between cells and background can vary so often according to the different, existing staining techniques, thickness of smear, microscope illumination and microscope's image acquisition procedure, as shown in fig. \ref{fig:images_types}. A standardization of the procedure should be really useful to avoid superfluous differences in similar images' features and to have fair comparisons among the several proposed methods. The main efforts towards the realization of a fully automatic blood cells segmentation and classification system cannot leave this aspect out. Usually, any  pattern recognition system is evaluated by performing tests using real set of samples. In case of diagnosis this is a key aspect. Any system for diagnosis should be tested on different images, acquired by different sources, and different specimens. This is crucial to guarantee the diagnostic capability and usefulness of a computer vision system.

Mathematical morphology techniques have been widely used for image processing purposes. Among the application fields, it has been applied for fingerprint feature extraction, recognition of handwritten digits, license plate detection, border extraction, denoising using morphological filters, text extraction and so on. Apart from this kind of fields, mathematical morphology has been employed successfully in biomedical image analysis, especially in preprocessing and segmentation techniques.

@@

%=====================================
% References, variant B: external bibliography
%=====================================
\reftitle{References}
\externalbibliography{yes}
\bibliography{bibliography}
